### 2008-03-21 trunk/Papers/dl-gre-08/bisim.tex
removed simulations

\section{Description Logics and Simulation} \label{sec:bisim}

As we mentioned, in this paper we will represent referring expressions
(REs) as formulas of description logic~\cite{baad:desc03}.  In order
to make this point, we will now first define the two description
logics we will be working with -- \alc\ and \el.

\begin{definition}
Formulas $\varphi$ of $\alc$ are generated by the following grammar:
$$
\varphi,\varphi' ::= p \mid \neg \varphi \mid \varphi \sqcap \varphi'
\mid \exists R. \varphi
$$
where $p$ is in the set of propositional symbols \prop, $R$ is in the
set of relational symbols \rel. $\el$ is the negation-free fragment of $\alc$.

Formulas of both $\alc$ and $\el$ are interpreted in ordinary
relational first-order models $\gM = (\Delta^\gM,|\cdot|^\gM)$ where
$\Delta^\gM$ is a non-empy set and $|\cdot|^\gM$ is an
interpretation function such that:
$$
\begin{array}{ccl}
|p|^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
|R|^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
|\neg \varphi|^\gM & = & \Delta^\gM \backslash |\varphi|^\gM\\
|\varphi \sqcap \varphi'|^\gM & = & |\varphi|^\gM \cap |\varphi'|^\gM\\
|\exists R.\varphi|^\gM & = & \{i \mid \mbox{for some } i', (i,i') \in |R|^\gM\\
& & \mbox{ and } i' \in |\varphi|^\gM \}.\\
\end{array}
$$

Given a model $\gM$ and an individual $i$ in $\gM$, let
$\propm(i) = \{p \in \prop \mid i \in |p|^\gM\}$.
\end{definition}

@@

@@

\begin{definition}
Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, and two individuals $i$ and $i'$ in $\gM$, an \emph{$\alc$-simulation} linking $i$ to $i'$  is a relation $Z
\subseteq \Delta^\gM \times \Delta^\gM$ such that 
\begin{itemize}
\item[i)] $(i,i') \in Z$.\\[-1.5em]
\item[ii)] if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.\\[-1.5em]
\item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
$(e_1,e_1') \in |R|^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.\\[-1.5em]
\item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
$(e_2,e_2') \in |R|^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.
\end{itemize}
%A maximal subset of $\Delta^\gM$ all of whose elements are pairwise
%bisimilar is called an \emph{\alc-bisimulation class} of $\gM$.
\end{definition}

We will say that an individual $i$ is \emph{\alc-similar} to $i'$
in a given model $\gM$ if for any formula $\varphi \in \alc$ $i \in \interp{\varphi}$ then
$i' \in \interp{\varphi}$.  Notice that, given that the language includes negation,
the notion of $\alc$-similarity is symmetric: if $i$ is \alc-similar to $i'$ then
$i'$ is \alc-similar to $i$.  This is not the case for all description
logic.  In particular, the notion of $\el$-similarity is not symmetric.
Given a model $\gM$ and an individual $i$ in $\gM$, we will write \simm$_\alc(i)$, the similarity set, to all the elements of $\gM$ which are \alc-similar to $i$.x

@@

>>>>

\section{Description logics and similarity}
\label{sec:bisim}

In this paper, we will represent referring expressions (REs) as
formulas of description logic~\cite{baad:desc03}.  In order to make
this point, we will now first define the two description logics we
will be working with: \alc\ and \el.

\begin{definition}
  Formulas (or concepts) $\varphi$ of $\alc$ are generated by the
  following grammar:
$$
\varphi,\varphi' ::= p \mid \neg \varphi \mid \varphi \sqcap \varphi'
\mid \exists R. \varphi
$$
where $p$ is in the set of propositional symbols \prop, $R$ is in the
set of relational symbols \rel. $\el$ is the negation-free fragment of $\alc$.

Formulas of both $\alc$ and $\el$ are interpreted in ordinary
relational first-order models $\gM = (\Delta^\gM,\interp{\cdot}^\gM)$ where
$\Delta^\gM$ is a non-empy set and $\interp{\cdot}^\gM$ is an
interpretation function such that:
$$
\begin{array}{ccl}
\interp{p}^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
\interp{R}^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
\interp{\neg \varphi}^\gM & = & \Delta^\gM \backslash \interp{\varphi}^\gM\\
\interp{\varphi \sqcap \varphi'}^\gM & = & \interp{\varphi}^\gM \cap \interp{\varphi'}^\gM\\
\interp{\exists R.\varphi}^\gM & = & \{i \mid \mbox{for some } i', (i,i') \in \interp{R}^\gM\\
& & \mbox{ and } i' \in \interp{\varphi}^\gM \}.\\
\end{array}
$$

Given a model $\gM$ and an individual $i$ in $\gM$, let
$\propm(i) = \{p \in \prop \mid i \in \interp{p}^\gM\}$.
\end{definition}

@@

A key notion in description logics that we will use here is
\emph{similarity} \todo{citation?}.  For any description logic $\gL$,
we will say that an individual $i$ is \emph{\gL-similar} to $i'$ in a
given model $\gM$ if for any formula $\varphi \in \gL$ such that $i
\in \interp{\varphi}$, we also have $i' \in \interp{\varphi}$.
Intuitively, this means that there is no $\gL$-formula that would
distinguish $i$ from $i'$.  We say that the \emph{\gL-similarity set}
of some individual $i$ is the set of all individuals to which $i$ is
\gL-similar.

Notice that similarity is not necessarily a symmetrical relation: For
instance, $f_1$ is \el-similar to $f_2$ in
Fig.~\ref{fig:dale-haddock}b, but $f_2$ is not \el-similar to $f_1$
(it satisfies the formula $\exists \mathsf{in}.\mathsf{hat}$ and $f_1$
doesn't).  However, \alc-similarity is a symmetrical relation because
\alc\ contains negation; and indeed, $f_1$ is not \alc-similar to
$f_2$ either because it satisfied $\neg \exists
\mathsf{in}.\mathsf{hat}$.  Because \alc\ is more expressive than \el,
it is generally possible for some individual $a$ to be \el-similar but
not \alc-similar to some individual $b$, but not vice versa.

\todo{need to cite \cite{blac:moda01,kurt:expr99} in the appropriate
  places}

\ignore{

Notice
that, given that the language includes negation, the notion of
$\alc$-similarity is symmetric: if $i$ is \alc-similar to $i'$ then
$i'$ is \alc-similar to $i$.  This is not the case for all description
logic.  In particular, the notion of $\el$-similarity is not
symmetric.  Given a model $\gM$ and an individual $i$ in $\gM$, we
will write \simm$_\alc(i)$, the similarity set, to all the elements of
$\gM$ which are \alc-similar to $i$.x

@@

\begin{definition}
Given a model $\gM = (\Delta^\gM,\interp{\cdot}^\gM)$, and two individuals $i$ and $i'$ in $\gM$, an \emph{$\alc$-simulation} linking $i$ to $i'$  is a relation $Z
\subseteq \Delta^\gM \times \Delta^\gM$ such that 
\begin{itemize}
\item[i)] $(i,i') \in Z$.\\[-1.5em]
\item[ii)] if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.\\[-1.5em]
\item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
$(e_1,e_1') \in \interp{R}^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in \interp{R}^\gM$ and
$(e_1',e_2') \in Z$.\\[-1.5em]
\item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
$(e_2,e_2') \in \interp{R}^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in \interp{R}^\gM$ and
$(e_1',e_2') \in Z$.
\end{itemize}
%A maximal subset of $\Delta^\gM$ all of whose elements are pairwise
%bisimilar is called an \emph{\alc-bisimulation class} of $\gM$.
\end{definition}

@@

}

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex
removed simulations

\medskip
\noindent
{\small
\begin{center}
\begin{tabular}{ll} \hline
\multicolumn{2}{l}{
\textsc{$\gL$-GRE Problem}}\\ \hline
\ \ Input: & A model $\gM$ and a set $A \subseteq \Delta^\gM$.\\
\ \ Output: & A formula $\varphi \in \gL$ such that $|\varphi|^\gM = A$\\
& \hspace*{0.5cm} (if such a formula exists).\\ \hline
\end{tabular}
\end{center}}
\medskip

@@

In the rest of this section, we will present algorithms that compute
the similarity sets of a given model for \alc\ and \el, and for each
similarity set, a characteristic formula that denotes exactly this
set.  In the \alc\ case, we adapt a standard algorithm from the
literature for computing \emph{bisimulation classes}; we will then
further adapt this algorithm for \el.  In effect, both algorithms
compute REs for all individuals in some model at the same time -- very
efficiently and without any danger of infinite regress.

@@

Theorem~\ref{bisim} states that in \alc, similarity sets correspond
exactly to \emph{bisimulation classes}, i.e.\ maximal sets of
individuals that can be linked by simulations.  Bisimulation classes
have been studied extensively in the literature, and there are several
efficient algorithms for computing \alc-bisimulation classes
\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
However, these algorithms will only compute the similarity sets
themselves. Here we extend the \newcite{hopc:algo71} algorithm such
that it computes, for each set, also a formula that denotes exactly
this set.  We can then use these formulas as representations of the
REs.

@@

\While{for some $\varphi \in \RE, |\varphi|^\gM>1$}{
   \For{$\varphi \in \RE, R \in \rel$}{
         add$_\mathcal{L}(\exists R.\varphi,\RE)$
   }
   \If{made no changes to \RE}{
      exit
      }
}
\end{algorithm}

@@

Now we can use the split between the bowls to distinguish the cups in
the second iteration.  The result of this is $\RE = \{ \mathsf{floor},
\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup} \sqcap \exists
\mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{floor}), \mathsf{cup} \sqcap \exists
\mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{table}), \mathsf{table} \}$.

At this point, all formulas except $\mathsf{table}$ denote singletons,
and further iterations don't allow us to refine $\mathsf{table}$; so
the algorithm terminates.  Each formula with a singleton extension
$\{a\}$ is a unique description of $a$; for instance, $\mathsf{cup}
\sqcap \exists \mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{table})$ is only satisfied by $c_2$, so we may
refer to $c_2$ as ``the cup in the bowl on the table''.  Notice that
the algorithm didn't focus on any particular individual; it
simultaneously generated REs for all individuals except for the two
tables.

>>>>

\medskip
\noindent
{\small
\begin{center}
\begin{tabular}{ll} \hline
\multicolumn{2}{l}{
\textsc{$\gL$-GRE Problem}}\\ \hline
\ \ Input: & A model $\gM$ and a set $A \subseteq \Delta^\gM$.\\
\ \ Output: & A formula $\varphi \in \gL$ such that
$\interp{\varphi}^\gM = A$\\ 
& \hspace*{0.5cm} (if such a formula exists).\\ \hline
\end{tabular}
\end{center}}
\medskip

@@

In the rest of this section, we will present algorithms that compute
the similarity sets of a given model for \alc\ and \el, and for each
similarity set, a characteristic formula that denotes exactly this
set.  In the \alc\ case, we adapt a standard algorithm from the
literature for computing \emph{simulation classes}; we will then
further adapt this algorithm for \el.  In effect, both algorithms
compute REs for all individuals in some model at the same time -- very
efficiently and without any danger of infinite regress.

@@

It can be shown that for \alc, the similarity sets of a finite model
coincide exactly with the \emph{simulation classes} of this model
\cite{blac:moda01}.  We don't have the space here to define simulation
classes, but they have been studied extensively in the literature, and
there are several efficient algorithms for computing \alc-simulation
classes
\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
However, these algorithms will only compute the simulation classes
themselves. Here we extend the \newcite{hopc:algo71} algorithm such
that it computes, along with each set, also a formula that denotes
exactly this set.  We can then use these formulas as representations
of the REs.

@@

\While{exists some $\varphi \in \RE, |\interp{\varphi}|^\gM>1$}{
   \For{$\varphi \in \RE, R \in \rel$}{
         add$_\mathcal{L}(\exists R.\varphi,\RE)$
   }
   \If{made no changes to \RE}{
      exit
      }
}
\end{algorithm}

@@

Now we can use the split between the bowls to distinguish the cups in
the second iteration.  The result of this is $\RE = \{ \mathsf{floor},
\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup} \sqcap \exists
\mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{floor}), \mathsf{cup} \sqcap \exists
\mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{table}), \mathsf{table} \}$.  At this point, all
formulas except $\mathsf{table}$ denote singletons, and further
iterations don't allow us to refine $\mathsf{table}$; so the algorithm
terminates.  Each formula with a singleton extension $\{a\}$ is a
unique description of $a$; for instance, $\mathsf{cup} \sqcap \exists
\mathsf{in}. (\mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{table})$ is only satisfied by $c_2$, so we may
refer to $c_2$ as ``the cup in the bowl on the table''.  Notice that
the algorithm didn't focus on any particular individual; it
simultaneously generated REs for all individuals except for the two
tables (which are similar to each other).
